Light inextensible strings (threads) under tension in the Schwarzschild geometry: a new interpretation for ``straight line'' in general relativity

### 2020-10-19 string_physics_teacher_revision.tex
further work for #20

\begin{document}
\begin{frontmatter}
\title{Light inextensible strings under tension in the Schwarzschild geometry}

@@

Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.

@@

The vacuum Einstein field equations for a static, spherically
symmetric mass are solved by the Schwarzschild
metric~\cite{schwarzschild1916}.  In units in which~$G=c=1$, this is:

@@

We also see a failure of classical geometry: the coordinate $r$ has
the defining feature that the circumference of a circle of radius $r$
has circumference $2\pi r$; but radial distances are not Euclidean.
To investigate this phenomenon, we consider the spatial
component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
throughout.

Students might need to be reminded of the concepts of proper time and
proper length: these are quantities measured {\em locally}.  In
particular, we note that proper length of an object is the length as
measured by a local observer at rest with respect to the object.  The
proper length of a string is the length measured by an inchworm as it
crawls along the string.  A circular string at constant~$r$ will,
trivially, have proper length~$2\pi r$.  Considering first the radial
component we may calculate the proper length $L\left(r_1,r_2\right)$
of a string stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$
at constant $\theta,\phi$:

@@

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial
strings has some non-Euclidean features.  Firstly, consider the proper
length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
Then

@@

\begin{enumerate}
\item null geodesics: paths for which $ds^2=0$ at all points.  Light
  travels on null geodesics and is arguably the ``straightest'' of
  straight lines.
\item timelike geodesics: paths for which $ds^2>0$ everywhere, and for
  which $\int\sqrt{ds^2}$ is a minimum.  These are the paths that
  objects in freefall follow.
\item taut string: paths for which $dt=0$, thereby considering the
  configuration of a string.  If the proper length
  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut.
\end{enumerate}

@@

\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  At this point it might be worth ``reminding''
students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$, and that equation~\ref{rdashdash}
reduces to this in the limit $r/2M\longrightarrow\infty$.

@@

Of course, Euler's method is numerically unstable and inefficient;
here use the R programming language~\cite{rcore2019} and the
\verb+deSolve+ package for ordinary differential
equations~\cite{soetart2010} using the standard \verb+lsoda+ solver.
The equations typically imply
$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C}=\infty$ for some
critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for flat
  space] and this makes numerical analysis challenging.

@@

Further analysis reveals something unexpected: for some values of
initial conditions, the string self-intersects even though it is under
tension everywhere and cannot support any nonzero bending moment.
Figure~\ref{closest_approach_self_intersecting} shows some
self-interesecting strings (Figure~\ref{light_closest_approach} shows
equivalent null geodesics for comparison).  A bisection technique
reveals that if the closest approach is greater than a critical value
of about~$\simeq 1.0761\cdot 2M$, the string does not self-intersect;
if the closest approach is less than this value, then the string
crosses itself at~$\phi=\pi/2$.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Using numerical methods to compute string
configuration is suitable for an undergraduate homework problem; the
difficult part is specifying sensible initial conditions.

The code producing the figures is available at github,
\url{http://github.com/RobinHankin/string}.

@@

\begin{figure}[h!] % Created by "flamm.R"
\centering
\includegraphics[width=\linewidth]{flamm_string.pdf}
\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
  path corresponding to a taut, light string}
\label{flamm}
\end{figure}

>>>>

\begin{document}
\begin{frontmatter}
\title{Light inextensible strings (threads) under tension in the
  Schwarzschild geometry: a new interpretation for ``straight line''
  in general relativity}

@@

Light inextensible string (thread) under tension is a stalwart feature
of elementary physics.  Here I show how considering such string in the
vicinity of a black hole, with the help of computer algebra systems,
can generate insight into the Schwarzschild geometry in the context of
an undergraduate homework problem, specifically by re-interpreting the
concept ``straight line''.  Light taut strings minimize their proper
length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some interesting
nonlocal phenomena: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.

@@

The vacuum Einstein field equations for a static, spherically
symmetric mass are solved by the Schwarzschild
metric~\cite{schwarzschild1916}.  In geometrized units, that is, a
system in which~$G=c=1$, the Schwarzschild metric is:

@@

We also see a failure of classical geometry: the coordinate $r$ has
the defining feature that the circumference of a circle of radius $r$
has circumference $2\pi r$; but radial distances are not Euclidean.
This is frequently expressed by saying that space is ``warped'', a
concept that has proved troublesome for generations of students.
Considering taut strings surely furnishes students with a useful
conceptual tool to understand the non-Euclidean geometry near a black
hole.  To investigate this phenomenon, we consider the spatial
component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$ of the
Schwarzschild metric, confined to equatorial points~($\theta=\pi/2$);
we understand~$r\geqslant 2M$ throughout.

Students might need to be reminded of the concepts of proper time and
proper length: these are quantities measured {\em locally}.  In
particular, we note that proper length of an object is the length as
measured by a local observer at rest with respect to the object.  The
proper length of a string is the length measured by an inchworm as it
crawls along the string.  A circular string at constant~$r$ will,
trivially, have proper length~$2\pi r$ [this follows from considering
  the Schwarzschild metric at constant $t,r,\theta$; the string will
  then have proper (inchworm) length $\int_{\phi=0}^{2\pi}ds=2\pi r$].
Considering now the radial component we may calculate the proper
length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
$\theta,\phi$:

@@

(students might need to be told explicitly that the right hand side of
equation~\ref{radial_string_length} not being equal to its Euclidean
value of $r_2-r_1$ is a direct illustration of the non-flatness of
space near a black hole).  Observe that the integration may be
performed with a lower limit of~$r_1=2M$ with no difficulty, even
though the radial component of the Schwarzschild metric approaches
infinity there.

One criticism of this analysis from a pedagogical perspective is that
the algebra is involved and nonintuitive.  However, we may apply
inchworm thinking via calculus and asymptotic series to show
simplified expressions for the proper length $\ell$ of radial strings.
Firstly, consider the proper length from $r=2M$ to $r=2M(1+w)$ where
$w>0, \left|w\right|\ll 1$.  Then

@@

\begin{enumerate}
\item null geodesics: paths for which $ds^2=0$ at all points.  Light
  travels on null geodesics and is arguably the ``straightest'' of
  straight lines.
\item timelike geodesics: paths for which $ds^2>0$ everywhere, and for
  which $\int\sqrt{ds^2}$ is a minimum.  These are the paths that
  objects in freefall follow.
\item taut string: paths for which $dt=0$, thereby considering the
  configuration of a string.  If the proper length
  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut in
  the sense that its inchworm-measured length is a minimum.
\end{enumerate}

@@

\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  At this point it might be worth reminding
students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$, and that equation~\ref{rdashdash}
reduces to this in the limit $r/2M\longrightarrow\infty$.

@@

Of course, Euler's method is numerically unstable and inefficient.
Here we use the R programming language~\cite{rcore2019} and the
\verb+deSolve+ package for ordinary differential
equations~\cite{soetart2010} using the standard \verb+lsoda+ solver.
The equations typically imply
$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C} r(\phi)=\infty$ for
some critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for
  flat space] and this makes numerical analysis challenging.

@@

Further analysis reveals something unexpected: for some values of
initial conditions, the string self-intersects even though it is under
tension everywhere and cannot support any nonzero bending moment.  One
might imagine our inchworm confidently declaring that the string was
everywhere ``straight'' in the sense of being a minimal proper length
path, and finding to his astonishment that he has traversed a closed
loop.

Figure~\ref{closest_approach_self_intersecting} shows some
self-interesecting strings (Figure~\ref{light_closest_approach} shows
equivalent null geodesics for comparison).  A bisection technique
reveals that if the closest approach is greater than a critical value
of about~$\simeq 1.0761\cdot 2M$, the string does not self-intersect;
if the closest approach is less than this value, then the string
crosses itself at~$\phi=\pi/2$.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  The device of using an inchworm to measure
proper distance along a non-extensible string serves to aid
understanding of the non-Euclidean nature of space near a black hole.

Using numerical methods to compute string configuration is suitable
for an undergraduate homework problem; the difficult part is
specifying sensible initial conditions.  The code producing the
figures is available at github,
\url{http://github.com/RobinHankin/string}.

@@

\begin{figure}[h!] % Created by "flamm.R"
\centering
\includegraphics[width=\linewidth]{flamm_string.pdf}
\caption{Perspective view of Flamm's paraboloid with superimposed
  minimal-length path corresponding to a taut, light string.  Proper
  length along this string corresponds to the path integral in
  $\mathbb{R}^3$.  Students might not be familiar with the fact that
  the vertical axis has no physical interpretation: there is no
  meaningful sense in which the spatial component of the Schwarzschild
  metric is a submanifold of a higher-dimensional Euclidean space}
\label{flamm}
\end{figure}
